Working Title: Roles and Expectations in Software Organizations

### 2011-09-27 soulgem.tex
Third draft, abstract.

Software professionals divide their work by roles, but the nature of these roles has not been examined nor questioned in our field. In this paper we formulate a theoretical framework to address this issue. The framework is based on sociological and organizational studies, and defines roles as sets of expectations placed on the members of the software organization.

The paper reports on an empirical case study of two software organizations, which was informed and inspired by the theoretical framework. In our study we found that roles were inconsistent within and between the organizations we studied, and that role definitions are shaped by institutional, historical, structural, and personal factors. Furthermore, role definitions evolve through organizational growth and through mismatches between expectations and performance, and people enacting different roles use different media to communicate.

Our contributions should help researchers to frame better studies of division of labour in software groups, and to avoid the pitfalls that come from unwarranted abstractions of software professional roles. They also provide practitioners with a conceptual tool to reason about social structure and organizational design.

>>>>

Software professionals divide their work by roles, but to date, researchers have not examined nor questioned the nature of these roles. To address this, we formulate a theoretical framework based on sociological and organizational research, in which roles are defined as sets of expectations placed on the members of the software organization.

The paper also reports on an empirical case study of two software organizations, which was informed and inspired by this theoretical framework. In our study we found that roles were inconsistent within and between the organizations we studied, and that role definitions are shaped by institutional, historical, structural, and personal factors. Furthermore, role definitions evolve through organizational growth and through mismatches between expectations and performance, and people enacting different roles vary in their selection of communication media.

Our contributions should help researchers to frame better studies of division of labour in software groups and to avoid unwarranted abstractions of software professional roles. They also provide practitioners with a conceptual tool to reason about social structure and organizational design.
